Stage-Splitting a Modal Language

### 2014-10-06 paper/paper.tex
more code colors

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}

@@

\begin{document}

\lstset{language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        moredelim=*[is][\color{red}]{1`}{`},
        moredelim=*[is][\color{blue}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

>>>>

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}

@@

\definecolor{stageOneColor}{RGB}{120,30,0}
\definecolor{stageTwoColor}{RGB}{0,0,127}
\begin{document}

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        moredelim=*[is][\color{stageOneColor}]{1`}{`},
        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

### 2014-10-06 paper/semantics.tex
more code colors

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
2`prev{`
  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
}`
\end{lstlisting}
The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
prev{
  next{ 
    prev{next {2`f 20`}} 2`*` prev{next {2`f 20`}} 2`* 7` 
  }
}
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
2`(f 20) * (f 20) * 7`
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
  1`let x = (`next{2`yhat`}1`, 7) in`
  next{prev{1`#1 x`} 2`*` prev{1`#1 x`} 2`*`  hold{1`#2 x`}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
  next{
    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*` 
    prev{1`#1 (`next {2`yhat`}1`, 7)`} 2`*`
    hold{1`#2 (`next {2`yhat`}1`, 7)`}
  }
}
\end{lstlisting}
We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
    next{ 2`yhat * yhat * 7` }
}
\end{lstlisting}
Once again, we lift the results of speculation into a substitution:
\begin{lstlisting} 
prev {
[2`yhat`|->2`f 20`]
[2`zhat`|->2`yhat * yhat * 7`]
    next{ 2`zhat` }
}
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  2`if true 
  then `hold{1`3+4`} 
  2`else` prev{1`spin() (* loops forever *)`}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

>>>>

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
2`prev{`
  1`next{` 2`prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * prev{`1`#1 (next {`2`f 20`1`}, 7)`2`} 
      * hold{`1`#2 (next {`2`f 20`1`}, 7)`2`} `1`}`2`
}`
\end{lstlisting}
The outermost $\next$ here is not yet a value, because there remains stage~\bbone\ work to do in its body.
So to proceed, the semantics must first search the body of the $\next$ for stage~\bbone\ subterms and reduce them in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
2`prev{`1`
  next{`2` 
    prev{`1`next {`2`f 20`1`}`2`} * prev{`1`next {`2`f 20`1`}`2`} * 7 
  `1`}`2`
}`
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
2`(f 20) * (f 20) * 7`
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

@@

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`let x = (next{`2`yhat`1`}, 7) in
  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
}`
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`next{`2`
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
  `1`}`2`
}`
\end{lstlisting}
We then speculate to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
    `1`next{`2`yhat * yhat * 7`1`}
`2`}`
\end{lstlisting}
Once again, we lift the results of speculation into a substitution:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
[zhat|->yhat * yhat * 7]
    `1`next{`2`zhat`1`}`2`
}`
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
1`next{
  `2`if true 
  then hold{`1`3+4`2`} 
  else prev{`1`spin() (* loops forever *)`2`}`1`
}`
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.
